EXAM Fits and Predictions

### 2023-11-14 Simulations/htw_exam.qmd
update with htw modelling

---
title: EXAM Fits and Predictions
date: last-modified
categories: [Simulation, ALM, EXAM, R]
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: true
---

@@

|                    | ALM Response Generation                                            |                                                                                               |
|------------------|------------------------------|-------------------------|
| Input Activation   | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$    | Input nodes activate as a function of Gaussian similarity to stimulus                         |
| Output Activation  | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$                        | Output unit $O_j$ activation is the weighted sum of input activations and association weights |
| Output Probability | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$                    | The response, $Y_j$ probabilites computed via Luce's choice rule                              |
| Mean Output        | $m(X) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$ | Weighted average of probabilities determines response to X                                    |
|                    | **ALM Learning**                                                   |                                                                                               |
| Feedback           | $f_j(Z) = e^{-c(Z-Y_j)^2}$                                         | feedback signal Z computed as similarity between ideal response and observed response         |
| magnitude of error | $\Delta_{ji}=(f_{j}(Z)-o_{j}(X))a_{i}(X)$                          | Delta rule to update weights.                                                                 |
| Update Weights     | $w_{ji}^{new}=w_{ji}+\eta\Delta_{ji}$                              | Updates scaled by learning rate parameter $\eta$.                                             |
|                    | **EXAM Extrapolation**                                             |                                                                                               |
| Instance Retrieval | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$                    | Novel test stimulus $X$ activates input nodes $X_i$                                           |
| Slope Computation  | $S =$ $\frac{m(X_{1})-m(X_{2})}{X_{1}-X_{2}}$                      | Slope value, $S$ computed from nearest training instances                                              |
| Response           | $E[Y|X_i] = m(X_i) + S \cdot [X - X_i]$                          | ALM response $m(X_i)$ adjusted by slope.                                                      |

@@

```{r}
#| label: tbl-e1-cogmodel
#| column: body-outset-right
#| tbl-cap: Fit Parameters and Model RMSE. The Test_RMSE column is the main performance indicator of interest, and represents the RMSE for just the testing data. The Fit_Method column indicates the data used to fit the model. Hybrid Model - w parameter. The $w$ parameter determines the balance between the ALM and EXAM response generation processes, and is only included for the hybrid model. A weight of .5 would indicate equal contribution from both models. $w$ values approaching 1 indicate stronger weight for EXAM.

reshaped_df <- all_combined_params %>%
  select(-Value) |>
  rename("Fit Method" = Fit_Method, "Test RMSE" = Test_RMSE) |>
  pivot_longer(cols=c(c,lr,w,"Test RMSE"),names_to="Parameter") |>
  unite(Group, Group, Parameter) %>%
  pivot_wider(names_from = Group, values_from = value)

header_df <- data.frame(
  col_keys = c("Model", "Fit Method","Constant_c", "Constant_lr", "Constant_w", "Constant_Test RMSE", "Varied_c", "Varied_lr", "Varied_w", "Varied_Test RMSE" 
               ),
  line1 = c("", "", "Constant", "", "", "", "Varied", "", "", ""),
  line2 = c("Model", "Fit Method", "c", "lr", "w", "Test RMSE", "c", "lr", "w", "Test RMSE")
)

ft <- flextable(reshaped_df) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% add_header_lines(values = " ") %>%
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>% 
  hline(part = "header", i = 2, j=3:6) %>% 
  hline(part = "header", i = 2, j=7:10)

@@

```{r}
#| label: tbl-e1-meanPreds
#| tbl-cap: Model Perforamnce - averaged over all X values/Bands. MAE=Mean Averaged Error, RMSE = Root mean squared error.

allPreds |> summarise(Error=mean(Error), Abs_Error=sqrt(mean(Abs_Error))) |> 
  mutate(Fit_Method=factor(Fit_Method, levels=c("Test Only", "Test & Train", "Train Only"))) |>
  tabulator(rows=c("Fit_Method", "Model"), columns=c("Group"), 
             `MAE` = as_paragraph(Error), 
            `RMSE` = as_paragraph(Abs_Error)) |> as_flextable()
```

## Varied Testing Predictions
```{r}
#| label: fig-model-preds-varied
#| fig-cap: Varied Group - Mean Model predictions vs. observations
#| fig-height: 12
#| fig-width: 14
#| column: screen-inset-right

@@

```{r fig.height=11, fig.width=11}
#| label: tbl-e1-predsV
#| tbl-cap: Varied group - mean model predictions vs. observations
#| column: screen-inset-right

@@

ft <- flextable(long_vPreds) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>%
  hline(part = "header", i = 1, j=4:9) %>%
  vline(j=c("Observed","ALM_Residual","EXAM_Residual")) %>%
  hline(part = "body", i=c(6,12)) |> 
  bold(i=long_vPreds$x %in% c(100,350,600), j=2)

@@

```{r}
#| label: tbl-e1-predsC
#| tbl-cap: Constant group - mean model predictions vs. observations
#| column: screen-inset-right

@@

ft <- flextable(long_cPreds) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>%
  hline(part = "header", i = 1, j=4:9) %>%
  vline(j=c("Observed","ALM_Residual","EXAM_Residual")) %>%
  hline(part = "body", i=c(6,12)) |> 
  bold(i=long_cPreds$x %in% c(100,350,600, 1000,1200), j=2)

@@

```{r}
#| label: tbl-e1-predsC2
#| tbl-cap: Constant group - mean model predictions vs. observations
#| eval: false 
tcte<- pluck(a_te_c, "test") |> rename(ALM=pred,Observed=y) %>% 
  cbind(.,EXAM=pluck(ex0_te_c, "test") |> pull(pred)) %>%
  cbind(., Hybrid=pluck(hybrid_te_c, "test") |> pull(pred))

tctetr<-pluck(a_tetr_c, "test") |> rename(ALM=pred,Observed=y) %>% 
  cbind(.,EXAM=pluck(ex0_tetr_c, "test") |> pull(pred)) %>%
  cbind(., Hybrid=pluck(hybrid_tetr_c, "test") |> pull(pred))

tctr<- pluck(a_tr_c, "test") |> rename(ALM=pred,Observed=y) %>% 
  cbind(.,EXAM=pluck(ex0_tr_c, "test") |> pull(pred)) %>%
  cbind(., Hybrid=pluck(hybrid_tr_c, "test") |> pull(pred))

pander(tcte, caption="Constant fit to test only")
pander(tctetr,caption="Constant fit to train and test")
pander(tctr,caption="Constant fit to train only")

```{r}
#| eval: false
pluck(a_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  learn_curve_plot(tr, vx, Resp,facet_var=x, groupVec=Resp,nbins=8)

>>>>

---
# title: EXAM Fits and Predictions
date: last-modified
categories: [Simulation, ALM, EXAM, R]
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: true
---

@@

|                    | **ALM Response Generation**                                            |                                                                                               |
|------------------|------------------------------|-------------------------|
| Input Activation   | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$    | Input nodes activate as a function of Gaussian similarity to stimulus                         |
| Output Activation  | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$                        | Output unit $O_j$ activation is the weighted sum of input activations and association weights |
| Output Probability | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$                    | The response, $Y_j$ probabilites computed via Luce's choice rule                              |
| Mean Output        | $m(X) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$ | Weighted average of probabilities determines response to X                                    |
|                    | **ALM Learning**                                                   |                                                                                               |
| Feedback           | $f_j(Z) = e^{-c(Z-Y_j)^2}$                                         | feedback signal Z computed as similarity between ideal response and observed response         |
| magnitude of error | $\Delta_{ji}=(f_{j}(Z)-o_{j}(X))a_{i}(X)$                          | Delta rule to update weights.                                                                 |
| Update Weights     | $w_{ji}^{new}=w_{ji}+\eta\Delta_{ji}$                              | Updates scaled by learning rate parameter $\eta$.                                             |
|                    | **EXAM Extrapolation**                                             |                                                                                               |
| Instance Retrieval | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$                    | Novel test stimulus $X$ activates input nodes $X_i$                                           |
| Slope Computation  | $S =$ $\frac{m(X_{1})-m(X_{2})}{X_{1}-X_{2}}$                      | Slope value, $S$ computed from nearest training instances                                              |
| Response           | $E[Y|X_i] = m(X_i) + S \cdot [X - X_i]$                          | ALM response $m(X_i)$ adjusted by slope.                                                      |

@@

```{r}
#| label: tbl-e1-cogmodel
#| tbl-cap: Fit Parameters and Model RMSE. The Test_RMSE column is the main performance indicator of interest, and represents the RMSE for just the testing data. The Fit_Method column indicates the data used to fit the model. The $w$ parameter determines the balance between the ALM and EXAM response generation processes, and is only included for the hybrid model. A weight of .5 would indicate equal contribution from both models. $w$ values approaching 1 indicate stronger weight for EXAM. 
##| column: body-outset-right

reshaped_df <- all_combined_params %>%
  select(-Value,-Test_RMSE) |>
  rename("Fit Method" = Fit_Method) |>
  pivot_longer(cols=c(c,lr,w),names_to="Parameter") %>%
  unite(Group, Group, Parameter) %>%
  pivot_wider(names_from = Group, values_from = value)

header_df <- data.frame(
  col_keys = c("Model", "Fit Method","Constant_c", "Constant_lr", "Constant_w", "Varied_c", "Varied_lr", "Varied_w"),
  line1 = c("", "", "Constant", "", "", "Varied", "",""),
  line2 = c("Model", "Fit Method", "c", "lr", "w", "c", "lr", "w")
)

ft <- flextable(reshaped_df) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% add_header_lines(values = " ") %>%
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>% 
  hline(part = "header", i = 2, j=3:5) %>% 
  hline(part = "header", i = 2, j=6:8)

@@

```{r}
#| label: tbl-e1-meanPreds
#| tbl-cap: Model Perforamnce - averaged over all X values/Bands. ME=Mean Average Error, RMSE = Root mean squared error. 
#| warning: false

allPreds |> summarise(Error=mean(Error), Abs_Error=sqrt(mean(Abs_Error))) |> 
  mutate(Fit_Method=factor(Fit_Method, levels=c("Test Only", "Test & Train", "Train Only"))) |>
  tabulator(rows=c("Fit_Method", "Model"), columns=c("Group"), 
             `ME` = as_paragraph(Error), 
            `RMSE` = as_paragraph(Abs_Error)) |> as_flextable()
```

## Varied Testing Predictions
```{r}
#| label: fig-model-preds-varied
#| fig-cap: Varied Group - Mean Model predictions vs. observations
#| fig-height: 12
#| fig-width: 14
##| column: screen-inset-right

@@

```{r fig.height=11, fig.width=11}
#| label: tbl-e1-predsV
#| tbl-cap: Varied group - mean model predictions vs. observations. Extrapolation Bands are bolded. For each Modelling fitting and band combination, the model with the smallest residual is highlighted. Only the lower bound of each velocity band is shown (bands are all 200 units). 
##| column: screen-inset-right

@@

ft <- flextable(long_vPreds) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>%
  hline(part = "header", i = 1, j=4:9) %>%
  vline(j=c("Observed","ALM_Residual","EXAM_Residual")) %>%
  hline(part = "body", i=c(6,12)) |> 
  bold(i=long_vPreds$x %in% c(100,350,600), j=2)

@@

```{r}
#| label: tbl-e1-predsC
#| tbl-cap: Constant group - mean model predictions vs. observations. The X values of Extrapolation Bands are bolded. For each Modelling fitting and band combination, the model with the smallest residual is highlighted. Only the lower bound of each velocity band is shown (bands are all 200 units). 
##| column: screen-inset-right

@@

ft <- flextable(long_cPreds) %>% 
  set_header_df(
    mapping = header_df,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>%
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues() %>%
  hline(part = "header", i = 1, j=4:9) %>%
  vline(j=c("Observed","ALM_Residual","EXAM_Residual")) %>%
  hline(part = "body", i=c(6,12)) |> 
  bold(i=long_cPreds$x %in% c(100,350,600, 1000,1200), j=2)

@@

```{r}
#| eval: false
#| include: false
pluck(a_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  learn_curve_plot(tr, vx, Resp,facet_var=x, groupVec=Resp,nbins=8)
